Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs

### 2019-07-23 manuscript/manuscript.Rmd
Bryan's edits

---
title: "Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs"
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
  ^*2*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Ecology Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: 'Addressing anthropogenic impacts on aquatic ecosystems has long been the focus of lake management. Controlling phosphorus and nitrogen can mitigate the impacts of eutrophication, but to determine the effectiveness requires long-term datasets.  A recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE), found stable water quality in the Northeast and Midwestern regions of the United States; however, trends at smaller scales may be masked.  We address this by analyzing the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.  URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. These data, included in LAGOSNE, allow for comparison of water quality trends at regional and state extents. We assess for trends with z-scores (i.e. scaled anomalies) calculated on a per-station basis and examine yearly averages.  Temperature and chlorophyll *a* are increasing.  Total nitrogen (TN) shows a weak increasing trend driven by low years in the early 1990s.  Total phosphorus (TP) and the total nitrogen:total phosphorus ratio (TN:TP) are stable.  Applying the site-specific z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll *a*, TN, TP, and TN:TP.  In short, productivity in Rhode Island lakes and reservoirs is increasing, in spite of stable nutrient regimes. Although not causal, this analysis suggests an association between lake temperature and productivity.  Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends. '
---

@@

Aquatic ecosystems have been altered as the result of human activities modifying nutrient cycling on a global scale [@vitousek1997human; @finlay2013human; @filippelli2008global]. Because of their position in the landscape, lakes can function as integrators and sentinels for these anthropogenic effects [@williamson2008lakes; @schindler2009lakes]. Increasing nutrient inputs, particularly of N and P, derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].  This eutrophication suggests an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutrophic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate]. To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication]. These studies indicate that spatial differences and relationships between N, P, and chlorophyll *a* exist and that long-term studies are needed to identify trends at local, regional, and national scales (I'm confused about this sentence- not sure how to edit.  Let's discuss it). Maybe...  These studies indicate that relationships between N, P, and chlorophyll *a* exist and these relationships are spatially complex; long-term studies are needed to identify trends at local, regional, and national scales.

Attempts to explore these spatial dyanmics are available (my attempt at intro sentence for paragraph).  Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis. These analyses can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing] (I'm not sure how this NLA fits into this.  Again, let's chat). Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases[@stoddard2016continental]. Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking]. These trophic state models indicate that landscape variables (i.e. ecoregion, elevation, and latitude) are important and that regional trends exist. Lake-specific drivers are also important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance]. Despite these challenges, it is important to study at multiple spatial scales because emergent trends on regional or continental scales may not be evident when studying individual lakes [@cheruvelil2013multi; @lottig2014long]. (thought- Maybe this paragraph just needs a introduction sentence. )

@@

Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).   URI’s Watershed Watch is a citizen science program founded in the late 1980's that has built a robust collaboration between URI scientists and a vast network of non-expert volunteer monitors.  Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season.  These efforts have been ongoing in some waterbodies since 1988.  These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing]. Here specifically, URIWW data contributed to not only the larger regional study by Oliver et al. [@Oliver_2017], but also allowed us to examine the long-term trends in Rhode Island.

Thus, the goals of this study were to examine approximately 25 years of lake and reservoir data in Rhode Island and answer two questions.  First, what are the state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature. Second, are water quality trends in Rhode Island similar to regional trends in the Northeastern United states. Another focus of this paper was to apply existing methods for examining long-term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long-term trends. Finally, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOSNE project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].

@@

For this study, we combined a long-term dataset on water quality of lakes in Rhode Island with a trend analysis based on centered and scaled water quality values (i.e. z-scores) to find increasing or decreasing annual water quality trends.  Details are outlined below.

@@

In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less. For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.  For each of these parameters, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 timeframe.  The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for TN, `r length(total_p_10yr_sites)` samples for TP.  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.

Additionally, prior studies have looked at water quality trends across a larger region of the Northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].  The authors found little change in water quality trends across this region [@Oliver_2017] .  We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].  Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.

@@

Anomaly analysis works best with a single measure (e.g., temperature) or with multiple measurements that are on the same scale.  The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.  For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.  To standardize these values, we used the common approach of dividing each anomaly by the standard deviation for the reference period (e.g., [@jones1996calculating]).  The resultant value is commonly referred to as a z-score.  We used these z-scores to examine each water quality parameter for a trend over the time period of 1993 to 2016.   Furthermore, since we are interested in water quality trends over time at individual sites, z-scores were calculated over the reference period, 1993-2016, for each site.  We refer to this approach as the site-specific z-scores.

@@

1. For each site, calculate the annual means, producing a single mean value for each site and year. This step prevents bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].  The per site means across years are assumed to be independent.
2. Calculate the long-term reference mean and reference standard deviation for all sites. This results in a single long-term mean and standard deviation for each of the sites.
3. Calculate the z-score for each annual mean at each site by subtracting the reference mean and dividing by the reference standard deviation.
4. Summarize by calculating the mean z-score per year for the entire group of sites.  The resultant values are analyzed for a trend over time.

@@

Prior studies have shown relatively stable water quality in the lakes of the Northeastern United Sates [@Oliver_2017].  While the University of Rhode Island's Watershed Watch data were included in that regional study,  we were curious if regional trends were masking local trends in Rhode Island and thus, wanted to compare the trends at the regional scale to the trends at the state scale.  The analysis conducted by Oliver et al. [-@Oliver_2017] is a robust approach; however, to make direct comparisons between Rhode Island and the region,  we re-analyzed the same dataset used by Oliver et al. [-@Oliver_2017] but using the trend analysis approach outlined above.

@@

Our analysis indicates that even when nutrient regimes exhibit relative stability (i.e. neither increasing nor decreasing over time), increases in productivity, as measured by chlorophyll *a*, occur.  One possible explanation is the long-term warming of Rhode Island lakes and reservoirs as also indicated by our analysis.  Also, scale does indeed matter when trying to identify long-term water quality trends. Similar to the results of Oliver et al. [-@Oliver_2017, ] our analysis shows little increasing trend in productivity at the regional scale (e.g. Northeastern and Mid-western United States).  However, at the local scale of the state of Rhode Island, there is a clear increasing trend in productivity.

@@

As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.  These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [EQIP at USDA (QK will provide), ]).  Additionally, these results point to the fact that productivity is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [Boyer et al. 2002].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [Nelson & Palmer, 2007]. Differences in regional and state level trends are driven by complex processes.

In addition to the annualized trends we address with this study, there are other trends that may be of interest.  For example, what are the trends for water quality at finer temporal scales such as monthly or seasonal trends?  Anecdotal evidence in Rhode Island [SHOULD REFERENCE THIS SOMEHOW - let's ask Elizabeth] points to greater increases in temperature earlier and later in the growing season and suggests a lengthening of the growing season. Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.  Additionally, are trends influenced by the current state of a given waterbody?  For instance, are oligotrophic lakes showing stronger trends than eutorphic lakes or are all lakes showing similar trends regardless of current trophic status?  These questions are beyond the scope of this study, but all warrant further, careful investigation.

@@

There are several possible approaches. First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters.  Green infrastructure approaches are one way to possibly achieve both goals [see notes].  Additionally, within-lake approaches may also be used.  A promising approach is the restoration of freshwater mussels to waterbodies that historically had those species. Some studies have shown reductions in both nutrients and algal biomass [see notes].

@@

The analysis approach we used here, site-specific z-scores, is not a novel method, but using it to examine water quality trends is novel as we could find few examples of using it specifically for water quality trends [NEED TO DO THIS LIT SEARCH].  This approach, and in particular non-scaled anomalies, does have a long history in the analysis of trends in climate [@hansen2006global, hansen2010global, @jones1996calculating, @jones1999surface].  We simply built on these methods and adapted them for use with long-term water quality trends.  While other methods are valid and robust (e.g. [@Oliver_2017]), we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.  In addition, the site-specific z-scores are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).  Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.  Without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analyses would have been impossible.  Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long term environmental trends.

>>>>

---
title: "Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs"
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
  ^*2*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Ecology Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: 'Addressing anthropogenic impacts on aquatic ecosystems has long been the focus of lake management. Controlling phosphorus and nitrogen can mitigate the impacts of eutrophication, but to determine the effectiveness requires long-term datasets.  A recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE), found stable water quality in the Northeast and Midwestern regions of the United States; however, trends at finer scales may be masked.  We address this by analyzing the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.  URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. These data, included in LAGOSNE, allow for comparison of water quality trends at regional and state extents. We assess for trends with z-scores (i.e. scaled anomalies) calculated on a per-station basis and examine yearly averages.  Temperature and chlorophyll *a* are increasing.  Total nitrogen (TN) shows a weak increasing trend driven by low years in the early 1990s.  Total phosphorus (TP) and the total nitrogen:total phosphorus ratio (TN:TP) are stable.  Applying the site-specific z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll *a*, TN, TP, and TN:TP.  In short, productivity in Rhode Island lakes and reservoirs is increasing, in spite of stable nutrient regimes. Although not causal, this analysis suggests an association between lake temperature and productivity.  Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends. '
---

@@

Aquatic ecosystems have been altered as the result of human activities modifying nutrient cycling on a global scale [@vitousek1997human; @finlay2013human; @filippelli2008global]. Because of their position in the landscape, lakes can function as integrators and sentinels for these anthropogenic effects [@williamson2008lakes; @schindler2009lakes]. Increasing nutrient inputs, particularly of nitrogen (N) and phosphorus (P), derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].  This eutrophication suggests an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutrophic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate]. To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication]. These studies indicate that complex relationships between N, P, and chlorophyll *a* exist and vary spatially; long-term studies are needed to identify trends at local, regional, and national scales.

Attempts to explore these spatial dynamics are available (my attempt at intro sentence for paragraph).  Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis. These analyses can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing] (I'm not sure how this NLA fits into this.  Again, let's chat). Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases[@stoddard2016continental]. Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking]. These trophic state models indicate that landscape variables (i.e. ecoregion, elevation, and latitude) are important and that regional trends exist. Lake-specific drivers are also important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance]. Despite these challenges, it is important to study at multiple spatial scales because emergent trends on regional or continental scales may not be evident when studying individual lakes [@cheruvelil2013multi; @lottig2014long]. (thought- Maybe this paragraph just needs a introduction sentence. ) (B: why not just remove this paragraph completely)

@@

Examining long-term trends in Rhode Island lakes is possible because of data gathered by University of Rhode Island’s Watershed Watch (URIWW).   URI’s Watershed Watch is a citizen science program founded in the late 1980's that has built a robust collaboration between URI scientists and a vast network of non-expert volunteer monitors.  Volunteer monitors are trained to collect *in situ* data as well as whole water samples during the growing season.  These efforts have been ongoing in some waterbodies since 1988.  These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing]. Here specifically, URIWW data contributed to not only the larger regional study by Oliver et al. [@Oliver_2017], but also allowed us to examine the long-term trends in Rhode Island.

The goals of this study were to examine ~25 years of lake and reservoir data in Rhode Island and answer two questions.  First, what are the state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature. Second, are water quality trends in Rhode Island similar to regional trends in the Northeastern United states. Another focus of this paper was to apply existing methods for examining long-term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long-term trends. Finally, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOSNE project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].

@@

@@

We analyzed a subset of the URI data that were collected from 1990 to 2016 during the months of May to October at a depth of 2m or less. Only lakes with at least 10 years of data were included. For this study, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.  The final dataset included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for TN, `r length(total_p_10yr_sites)` samples for TP.  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.

Prior studies have looked at water quality trends across a larger region of the Northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].  The authors found little change in water quality trends across this region [@Oliver_2017] .  We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].  Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.

@@

Anomaly analysis works best with a single measure (e.g., temperature) or with multiple measurements that are on the same scale.  The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.  For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.  To standardize these values, we used the common approach of dividing each anomaly by the standard deviation for the reference period (e.g., [@jones1996calculating]).  The resultant value is commonly referred to as a z-score.  We used these z-scores to examine each water quality parameter for a trend over the time period of 1993 to 2016 (B: 1993 or 1990?).   Furthermore, since we are interested in water quality trends over time at individual sites, z-scores were calculated over the reference period, 1993-2016, for each site.  We refer to this approach as the site-specific z-scores.

@@

1. For each site, calculate the annual means, producing a single mean value for each site and year. This step prevents bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].  The per site means across years are assumed to be independent.
2. Calculate the long-term reference mean and reference standard deviation for all sites. This results in a single long-term mean and standard deviation for each of the sites. (B: i'm confused-the first sentence suggests you get single values for the mean and SD across all sites but the second sentence says the a single mean and SD are calculated for each site.  If it is the second then the index period will vary by site and this needs to be explained.)
3. Calculate the z-score for each annual mean at each site by subtracting the reference mean and dividing by the reference standard deviation.
4. Summarize by calculating the mean z-score per year for the entire group of sites.  The resultant values are analyzed for a trend over time.

@@

Oliver et al (2017) have shown relatively stable water quality in the lakes of the Northeastern United States; the University of Rhode Island's Watershed Watch data were included in this regional study. However, we hypothesized in the case of Rhode Island regional trends were masking local trends.  Therefore we decided to reanalyze the the LAGOSNE data to compare the trends at the regional scale to the trends at the Rhode Island state scale using the approach outlined above.

@@

Our analysis indicates that even when nutrient regimes exhibit relative stability (i.e. neither increasing nor decreasing over time), increases in productivity, as measured by chlorophyll *a*, occur.  Over the same period we also demonstrate long-term warming of RI lakes and reservoirs.  This suggests that the observered increase in producticity, as measured by chlorophyll *a* is a direct result of warming waters and not a response to changes in nutrient condition.  Also, scale does indeed matter when trying to identify long-term water quality trends. Similar to the results of Oliver et al. [-@Oliver_2017, ] our analysis shows little increasing trend in productivity at the regional scale (e.g. Northeastern and Mid-western United States).  However, at the local scale of the state of Rhode Island, there is a clear increasing trend in productivity.

@@

As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are relatively stable.  Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [EQIP at USDA (QK will provide), ]).  Additionally, these results point to the fact that productivity is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [Boyer et al. 2002].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [Nelson & Palmer, 2007]. Differences in regional and state level trends are driven by complex processes.

In addition to the annualized trends we address with this study, there are other trends that may be of interest.  For example, what are the trends for water quality at finer temporal scales such as monthly or seasonal trends?  Anecdotal evidence in Rhode Island [SHOULD REFERENCE THIS SOMEHOW - let's ask Elizabeth] points to greater increases in temperature earlier and later in the growing season and suggests a lengthening of the growing season. Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.  Additionally, are trends influenced by the current state of a given waterbody?  For instance, are oligotrophic lakes showing stronger trends than eutrophic lakes or are all lakes showing similar trends regardless of current trophic status?  These questions are beyond the scope of this study, but all warrant further, careful investigation.

@@

There are several possible approaches. First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters.  Green infrastructure approaches are one way to possibly achieve both goals [see notes].  Additionally, within-lake approaches may also be used.  A promising approach is the restoration of freshwater mussels to waterbodies that historically had those species. Some studies have shown reductions in both nutrients and algal biomass [see notes].  (B: where did this come from?  Are we sure mussels are this important?)

@@

The analysis approach we used here, site-specific z-scores, is not a novel method, but using it to examine water quality trends is a new application of the technique; we could find few examples of using it specifically for water quality trends [NEED TO DO THIS LIT SEARCH].  This approach, and in particular non-scaled anomalies, does have a long history in the analysis of trends in climate [@hansen2006global, hansen2010global, @jones1996calculating, @jones1999surface].  We simply built on these methods and adapted them for use with long-term water quality trends.  While other methods are valid and robust (e.g. [@Oliver_2017]), we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.  In addition, the site-specific z-scores are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).  Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.  Without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analyses would have been impossible.  Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long term environmental trends.
